Model to study co-evolution of trade an culture

### 2015-04-06 article.tex
added et al.

%XRC trade networks are increasingly being recognised as a key process in ancient Mediterranean
Trades are being increasingly recognised as a key process in the structure of ancient Mediterranean societies. For example, the use of table ware in the Roman empire  has been linked to the trade network~\cite{brughmans_connecting_2010}. Interestingly this trade network has been shown to be a market economy~\cite{temin_market_2001}. The work presented in this paper proposes a framework to further study the relation between trade and culture within the roman empire. This work takes place within the EPNET project~\cite{remesal_epnet_2014}.

@@

To the best of our knowledge two works have previously proposed frameworks that could be used to model the co-evolution of trade and culture~\cite{bentley_specialisation_2005,macmillan_agent-based_2008}. The first proposes a relatively simple model which is limited by 2 products, a fixed behaviour of the agents and a limited variation of cultural mechanisms. The second proposes a more complete model including a large number of goods (certain being essential and the others non-essential), but suffer from a greater complexity and the absence of the notion of trade networks. Moreover, these works do not compare the dynamics obtained to the previous studies on cultural change and trading mechanisms.

@@

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population~\cite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution~\cite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies~\cite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution~\cite{burger_absence_2007}, diffusion processes~\cite{fort_synthesis_2012}, phylogenetic analysis of material culture~\cite{obrien_cladistics_2001} or detection of rates of change in the archaeological record~\cite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
These analysis are often based on the theory of neutral evolution~\cite{neiman_stylistic_1995}, which explains variation in the distribution of traits as a function of their frequency distribution in the population. Thus, a trait is assumed to lack any selective bias and its transmission will depend on its popularity amongst the rest of individuals. The comparison of a given frequency distribution found in data against this unbiased transmission model is traditionally used as a null-hypothesis testing for selection. There exist a large number of works using this approach as the starting point to analyse patterns of cultural change in past societies~\cite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

Unbiased transmission generates a clear pattern of power law distributions that can be replicated with a simple random copying transmission mechanism~\cite{bentley_random_2004}. Within this framework, an individual will copy the traits of a randomly chosen individual with a given probability. This copy can potentially introduce some errors in the acquired trait, which account for innovation processes. The individual will in turn continue to spread these cultural traits which will be further adopted by other individuals. This basic model can be enriched by several additional processes both in the innovation~\cite{schillinger_copying_2014,sole_evolutionary_2013,ziman_technological_2003} and the transmissio~\cite{heyes_social_1994,henrich_evolution_2003}.

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\cite{temin_market_2001,temin_economy_2006}. 
Based on these conclusions, further works have attempt to define more exactly the characteristics of this market such, e.g. level of intervention of the state and its impact, or the relation-ship between local and long-distance trade~\cite{wilson_approaches_2009}.

Few works have attempt to model the interactions between economy and cultural change in the Roman empire~\cite{terpstra_trade_2011,kessler_organization_2007,scheidel_model_2007}. An analysis based on historical documents has revealed the importance of economical and social institutions for the efficiency of trades of grains in the early Roman Empire~\cite{kessler_organization_2007}. The same type of analysis has also revealed the importance of a number of key factors (such as commercial development, population growth and geographical mobility) for the growth of income and population in the Roman Italy~\cite{scheidel_model_2007}. Finally, a micro-economic model has determined the importance of reputation and threat of expulsion from a trade network as mechanisms to maintain the overall efficiency of the trade network~\cite{terpstra_trade_2011}. These works therefore reveal the strong relation between trade and cultural change in the Roman empire. However a large percentage of the datasets used for these studies are based on material objects found in archaeological contexts such as pottery. Most of these objects are usually represented following a neutral model, but the weight of this model within a market system is not studied.

@@

The Agent Based Models (ABM) methodology offers a way to produce this kind of model within a computer~\cite{lake_trends_2014}. An ABM is composed of a group of agents performing actions within a virtual environment. The environment and the behaviour of the agents, despite being an simpler than the real world, are designed to be analogous to their real-world counterpart. The experimenter hand designs the agents, the environment and their interaction, and observes the outcome of all interactions through time in order to compare it to the archaelogical data obtained from the real world. The goal of the experimenter is therefore to find a plausible set of assumptions used during the creation of the ABM to produce results close to what is observed in reality.

A first example of its potential is given through a model of the trade network of tableware in the roman east~\cite{brughmans_roman_2014}. Among the parameters tested, preliminary results of this model, show that the number of links between production site has a strong influence on the results obtained. A second more complete example proposes a detailed model of the prehispanic Pueblo societies~\cite{kohler_modelling_2012}. This project proposes an ABM integrating numerous effects such as food consumption, production and landscape. The questions studied are  numerous and some works are performed on the economical system and its ties with social transmissions~\cite{kobti_emergence_2006,cockburn_simulating_2013}. However the work done in this project consider a pre-market society in a specific village which makes it difficult to reach more general conclusions, notably applicable to the Roman Empire.

@@

The first set of implementations is designed to reproduce a scenario of unbiased transmission, where each product is a cultural trait without intrinsic positive or negative weight \cite{bentley_random_2004,bentley_specialisation_2005,mesoudi_random_2009}. 
Under this hypothesis, the processes of \emph{production} and \emph{trade} are not relevant, and as a consequence, they do not modify the content of the quantity vectors of the agents.

@@

For each of the setup 100 runs of 200000 timestep is performed. The parameters used are presented in table~\ref{tab:parameters}. Finally, remember that in our experiment one has to take in account that agents modifies their prices or exchange their price only every $culturalStep$ steps. Between, twos such steps, the agent exchange goods given their own prices. Therefore we choose to run our simulation during 10000 step which correspond the 1000 time steps used by \cite{bentley_random_2004,mesoudi_random_2009}.

@@

We observe that prices are indeed converging to the optimal prices which means that our innovation and imitation process are valid trade mechanisms. Notably, they produce results similar to the ones obtained in~\cite{gintis_emergence_2006}.

@@

In future works numerous, more realistic and complex dynamics can be studied in the same model. First the constraints on the production could be relaxed so that each agent can choose the products it wishes to produce at the current time step. Moreover, meterological factors can modify the quality of the production (noisy outcome of the process) thus making the process more realistic and more complex~\cite{bentley_specialisation_2005}.

@@

The trading theories studied also can be extended to the \emph{Prospect Theory}~\cite{kahneman_prospect_1979} which proposes to study more realistic trade mechanisms taking into account the biases of humans. Another economical framework we envision is the Agent based Computational Economy, ACE, which is  addressing economical problems raising from the interactions of a large number of agents~\cite{tesfatsion_introduction_2001}. On top of these trading theories the influence of the network of trades by itself can be studied. Various networks can be envisioned ranging from fully connected to small world, and taking into account possible delays or the sudden rupture of a link.

On the cultural transmission side, the vector value could be learned using any mechanism of ``social learning'' (as defined by \cite{lycett_cultural_2015}) known in the literature (teaching, different kind of copying mechanisms,\ldots) and/or integrate any cognitive/environmental bias that could be studied (see again \cite{lycett_cultural_2015} for some kind of bias that could be implemented). Agents could be also endowed with self-adaptation abilities implemented thanks to Reinforcement Learning or Developmental Learning. Another possible track to implement the self-adaptation abilities is to rely on the cognitive approaches that propose to model the complexity of the human behaviour in general representations such as BDI.

@@

The model was created using Pandora \cite{rubiocampillo_2014}. R was used for figures and statistical analysis \cite{rdev_2012}. The source code of the model is licensed under a GNU General Public License and will be available for download at publication.

>>>>

%XRC trade networks are increasingly being recognised as a key process in ancient Mediterranean
Trades are being increasingly recognised as a key process in the structure of ancient Mediterranean societies. For example, the use of table ware in the Roman empire  has been linked to the trade network~\shortcite{brughmans_connecting_2010}. Interestingly this trade network has been shown to be a market economy~\shortcite{temin_market_2001}. The work presented in this paper proposes a framework to further study the relation between trade and culture within the roman empire. This work takes place within the EPNET project~\shortcite{remesal_epnet_2014}.

@@

To the best of our knowledge two works have previously proposed frameworks that could be used to model the co-evolution of trade and culture~\shortcite{bentley_specialisation_2005,macmillan_agent-based_2008}. The first proposes a relatively simple model which is limited by 2 products, a fixed behaviour of the agents and a limited variation of cultural mechanisms. The second proposes a more complete model including a large number of goods (certain being essential and the others non-essential), but suffer from a greater complexity and the absence of the notion of trade networks. Moreover, these works do not compare the dynamics obtained to the previous studies on cultural change and trading mechanisms.

@@

The first aspect of our model is focused on the dynamics of cultural change. They comprise the collection of processes that promotes or inhibits the spread of information by social interaction within a population~\shortcite[3]{boyd_origin_2005}. An increasing number of social scientists are using an evolutionary framework to model these dynamics, thus fostering the development of transdisciplinary efforts designed to understand cultural evolution~\shortcite{henrich_evolution_2003}. Archaeology has a rich tradition of evolutionary studies~\shortcite{lycett_cultural_2015}, exploring concepts such as gene-culture coevolution~\shortcite{burger_absence_2007}, diffusion processes~\shortcite{fort_synthesis_2012}, phylogenetic analysis of material culture~\shortcite{obrien_cladistics_2001} or detection of rates of change in the archaeological record~\shortcite{premo_cultural_2014}.

A large percentage of these studies focus on the biases that affect the transmission of cultural traits, including the relevance of intrinsic traits and frequency dependence (e.g. conformism and anti-conformism). 
These analysis are often based on the theory of neutral evolution~\shortcite{neiman_stylistic_1995}, which explains variation in the distribution of traits as a function of their frequency distribution in the population. Thus, a trait is assumed to lack any selective bias and its transmission will depend on its popularity amongst the rest of individuals. The comparison of a given frequency distribution found in data against this unbiased transmission model is traditionally used as a null-hypothesis testing for selection. There exist a large number of works using this approach as the starting point to analyse patterns of cultural change in past societies~\shortcite{lipo_neutralitystyle_2001,shennan_ceramic_2001,steele_ceramic_2010,kandler_nonequilibrium_2013,porcic_exploring_2014,crema_approximate_2014}.

Unbiased transmission generates a clear pattern of power law distributions that can be replicated with a simple random copying transmission mechanism~\shortcite{bentley_random_2004}. Within this framework, an individual will copy the traits of a randomly chosen individual with a given probability. This copy can potentially introduce some errors in the acquired trait, which account for innovation processes. The individual will in turn continue to spread these cultural traits which will be further adopted by other individuals. This basic model can be enriched by several additional processes both in the innovation~\shortcite{schillinger_copying_2014,sole_evolutionary_2013,ziman_technological_2003} and the transmissio~\shortcite{heyes_social_1994,henrich_evolution_2003}.

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches~\shortcite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\shortcite{temin_market_2001,temin_economy_2006}. 
Based on these conclusions, further works have attempt to define more exactly the characteristics of this market such, e.g. level of intervention of the state and its impact, or the relation-ship between local and long-distance trade~\shortcite{wilson_approaches_2009}.

Few works have attempt to model the interactions between economy and cultural change in the Roman empire~\shortcite{terpstra_trade_2011,kessler_organization_2007,scheidel_model_2007}. An analysis based on historical documents has revealed the importance of economical and social institutions for the efficiency of trades of grains in the early Roman Empire~\shortcite{kessler_organization_2007}. The same type of analysis has also revealed the importance of a number of key factors (such as commercial development, population growth and geographical mobility) for the growth of income and population in the Roman Italy~\shortcite{scheidel_model_2007}. Finally, a micro-economic model has determined the importance of reputation and threat of expulsion from a trade network as mechanisms to maintain the overall efficiency of the trade network~\shortcite{terpstra_trade_2011}. These works therefore reveal the strong relation between trade and cultural change in the Roman empire. However a large percentage of the datasets used for these studies are based on material objects found in archaeological contexts such as pottery. Most of these objects are usually represented following a neutral model, but the weight of this model within a market system is not studied.

@@

The Agent Based Models (ABM) methodology offers a way to produce this kind of model within a computer~\shortcite{lake_trends_2014}. An ABM is composed of a group of agents performing actions within a virtual environment. The environment and the behaviour of the agents, despite being an simpler than the real world, are designed to be analogous to their real-world counterpart. The experimenter hand designs the agents, the environment and their interaction, and observes the outcome of all interactions through time in order to compare it to the archaelogical data obtained from the real world. The goal of the experimenter is therefore to find a plausible set of assumptions used during the creation of the ABM to produce results close to what is observed in reality.

A first example of its potential is given through a model of the trade network of tableware in the roman east~\shortcite{brughmans_roman_2014}. Among the parameters tested, preliminary results of this model, show that the number of links between production site has a strong influence on the results obtained. A second more complete example proposes a detailed model of the prehispanic Pueblo societies~\shortcite{kohler_modelling_2012}. This project proposes an ABM integrating numerous effects such as food consumption, production and landscape. The questions studied are  numerous and some works are performed on the economical system and its ties with social transmissions~\shortcite{kobti_emergence_2006,cockburn_simulating_2013}. However the work done in this project consider a pre-market society in a specific village which makes it difficult to reach more general conclusions, notably applicable to the Roman Empire.

@@

The first set of implementations is designed to reproduce a scenario of unbiased transmission, where each product is a cultural trait without intrinsic positive or negative weight \shortcite{bentley_random_2004,bentley_specialisation_2005,mesoudi_random_2009}. 
Under this hypothesis, the processes of \emph{production} and \emph{trade} are not relevant, and as a consequence, they do not modify the content of the quantity vectors of the agents.

@@

For each of the setup 100 runs of 200000 timestep is performed. The parameters used are presented in table~\ref{tab:parameters}. Finally, remember that in our experiment one has to take in account that agents modifies their prices or exchange their price only every $culturalStep$ steps. Between, twos such steps, the agent exchange goods given their own prices. Therefore we choose to run our simulation during 10000 step which correspond the 1000 time steps used by \shortcite{bentley_random_2004,mesoudi_random_2009}.

@@

We observe that prices are indeed converging to the optimal prices which means that our innovation and imitation process are valid trade mechanisms. Notably, they produce results similar to the ones obtained in~\shortcite{gintis_emergence_2006}.

@@

In future works numerous, more realistic and complex dynamics can be studied in the same model. First the constraints on the production could be relaxed so that each agent can choose the products it wishes to produce at the current time step. Moreover, meterological factors can modify the quality of the production (noisy outcome of the process) thus making the process more realistic and more complex~\shortcite{bentley_specialisation_2005}.

@@

The trading theories studied also can be extended to the \emph{Prospect Theory}~\shortcite{kahneman_prospect_1979} which proposes to study more realistic trade mechanisms taking into account the biases of humans. Another economical framework we envision is the Agent based Computational Economy, ACE, which is  addressing economical problems raising from the interactions of a large number of agents~\shortcite{tesfatsion_introduction_2001}. On top of these trading theories the influence of the network of trades by itself can be studied. Various networks can be envisioned ranging from fully connected to small world, and taking into account possible delays or the sudden rupture of a link.

On the cultural transmission side, the vector value could be learned using any mechanism of ``social learning'' (as defined by \shortcite{lycett_cultural_2015}) known in the literature (teaching, different kind of copying mechanisms,\ldots) and/or integrate any cognitive/environmental bias that could be studied (see again \shortcite{lycett_cultural_2015} for some kind of bias that could be implemented). Agents could be also endowed with self-adaptation abilities implemented thanks to Reinforcement Learning or Developmental Learning. Another possible track to implement the self-adaptation abilities is to rely on the cognitive approaches that propose to model the complexity of the human behaviour in general representations such as BDI.

@@

The model was created using Pandora \shortcite{rubiocampillo_2014}. R was used for figures and statistical analysis \shortcite{rdev_2012}. The source code of the model is licensed under a GNU General Public License and will be available for download at publication.
